---
title: "Assignment 2: Coding Basics"
author: "Angela Edwards"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. Naming a sequence of 1 to 100 with increments of 4, with the name hundred_sequence:
hundred_sequence <- seq(1, 100, 4)

#2. Finding the mean and median of hundred_ sequence:
mean(hundred_sequence)
median(hundred_sequence)

#3. Asking R to determine whether the mean of hundred_sequence is greater than its mean:
mean(hundred_sequence) > median(hundred_sequence)

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#5a & 6) Creating vector with four student names; this is a character vector.
student_name <- c("Kai", "Casey", "Pat", "Jessie")

#5b & 6) Creating vector with four test scores; this is an integer vector.
test_score <- c(88, 92, 49, 75)

#5c & 6) Creating vector for having passed the test; this is a logical vector.
passed_test <- c(TRUE, TRUE, FALSE, TRUE)

#7 & 8) Combining vectors into a data frame with an assigned name and named columns:
student_test_scores <- cbind(student_name, test_score, passed_test)

  

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:
In a matrix, all columns must have the same mode and the same length. In this data frame, the columns do have the same length, but they have different modes, specifically: character, integer, and logical.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
# Using the if and else statements:
passing_score_1 <- function(x){
  passed <- x>50
  failed <- x<50
  print(list(passed = passed,
        failed = failed))
}

passed_1 <- passing_score_1(test_score)
passed_1

# Using the ifelse statement:
passing_score_2 <- function(x){
  ifelse(x>50, print(TRUE), print(FALSE))
}

passed_2 <- passing_score_2(test_score)
passed_2

# Alternate print command using the ifelse statement:
passing_score_3 <- function(x){
  ifelse(x<50, print("Failed"), print("Passed"))
}

passed_3 <- passing_score_3(test_score)
passed_3

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:
I'm not sure if I did this exercise correctly, but it looks like they both worked, at least when using the integer vector of test_score. The different versions showed the information a bit differently; the first example displays output as two lists ($passed and $failed), whereas the two if/else examples show one line of output. 
