---
title: "Assignment 4: Data Wrangling"
author: "Angela Edwards"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```


## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions

1.  Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Feb 20th @ 5:00pm.

## Set up your session 

1a. Load the `tidyverse`,  `lubridate`, and `here` packages into your session.

1b. Check your working directory.

1c. Read in all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Apply the `glimpse()` function to reveal the dimensions, column names, and structure of each dataset. 

```{r, message = FALSE}
#1a Loading necessary packages.
library(tidyverse)
library(lubridate)
library(here)
library(formatR)

# library(knitr)
# library(plyr)

#1b Checking working directory.
getwd()

#1c Reading in the four EPAair datasets.
EPAair_O3_NC2018 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv", stringsAsFactors = TRUE)
EPAair_O3_NC2019 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv", stringsAsFactors = TRUE)
EPAair_PM25_NC2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv", stringsAsFactors = TRUE)
EPAair_PM25_NC2019 <- read.csv("./Data/Raw/EPAair_PM25_NC2019_raw.csv", stringsAsFactors = TRUE)


#2 Using glimpse to peruse each dataset.
glimpse(EPAair_O3_NC2018)
glimpse(EPAair_O3_NC2019)
glimpse(EPAair_PM25_NC2018)
glimpse(EPAair_PM25_NC2019)

```


## Wrangle individual datasets to create processed files.

3. Change date columns to be date objects.

4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE

5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).

6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3 Changing 'Date' into date object for all datasets.
EPAair_O3_NC2018$Date <- as.Date(EPAair_O3_NC2018$Date, format = "%m/%d/%Y")

EPAair_O3_NC2019$Date <- as.Date(EPAair_O3_NC2019$Date, format = "%m/%d/%Y")

EPAair_PM25_NC2018$Date <- as.Date(EPAair_PM25_NC2018$Date, format = "%m/%d/%Y")

EPAair_PM25_NC2019$Date <- as.Date(EPAair_PM25_NC2019$Date, format = "%m/%d/%Y")

# Creating unique row numbers to help with issues of duplication after subsetting.
EPAair_O3_NC2018 <- 
  EPAair_O3_NC2018 %>%
  dplyr::mutate(Row=row_number()) %>%
  relocate(Row)

EPAair_O3_NC2019 <- 
  EPAair_O3_NC2019 %>%
  dplyr::mutate(Row=row_number()) %>%
  relocate(Row)

EPAair_PM25_NC2018 <- 
  EPAair_PM25_NC2018 %>%
  dplyr::mutate(Row=row_number()) %>%
  relocate(Row)

EPAair_PM25_NC2019 <- 
  EPAair_PM25_NC2019 %>%
  dplyr::mutate(Row=row_number()) %>%
  relocate(Row)


#4 Using select to choose specific columns.
EPAair_O3_NC2018_subset <- select(EPAair_O3_NC2018, Row, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)

EPAair_O3_NC2019_subset <- select(EPAair_O3_NC2019, Row, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)

EPAair_PM25_NC2018_subset <- select(EPAair_PM25_NC2018, Row, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)

EPAair_PM25_NC2019_subset <- select(EPAair_PM25_NC2019, Row, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)

#5 For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5".

# First converting AQS_PARAMETER_DESC factor variable into character variable for both datasets in order to replace the values.
EPAair_PM25_NC2018_subset$AQS_PARAMETER_DESC <- as.character(EPAair_PM25_NC2018_subset$AQS_PARAMETER_DESC)

EPAair_PM25_NC2019_subset$AQS_PARAMETER_DESC <- as.character(EPAair_PM25_NC2019_subset$AQS_PARAMETER_DESC)

#Replacing current values with "PM2.5".
EPAair_PM25_NC2018_subset$AQS_PARAMETER_DESC[EPAair_PM25_NC2018_subset$AQS_PARAMETER_DESC == "Acceptable PM2.5 AQI & Speciation Mass" | EPAair_PM25_NC2018_subset$AQS_PARAMETER_DESC == "PM2.5 - Local Conditions"] <- "PM2.5"

EPAair_PM25_NC2019_subset$AQS_PARAMETER_DESC[EPAair_PM25_NC2019_subset$AQS_PARAMETER_DESC == "Acceptable PM2.5 AQI & Speciation Mass" | EPAair_PM25_NC2019_subset$AQS_PARAMETER_DESC == "PM2.5 - Local Conditions"] <- "PM2.5"

# Convert AQS_PARAMETER_DESC character variable back to factor to match up with the other datasets.
EPAair_PM25_NC2018_subset$AQS_PARAMETER_DESC <- as.factor(EPAair_PM25_NC2018_subset$AQS_PARAMETER_DESC)

EPAair_PM25_NC2019_subset$AQS_PARAMETER_DESC <- as.factor(EPAair_PM25_NC2019_subset$AQS_PARAMETER_DESC)


#6 Saving each dataset in the processed folder. 
write.csv(EPAair_O3_NC2018_subset, row.names = FALSE, file = "./Data/Processed/EPAair_O3_NC2018_processed.csv")

write.csv(EPAair_O3_NC2019_subset, row.names = FALSE, file = "./Data/Processed/EPAair_O3_NC2019_processed.csv")

write.csv(EPAair_PM25_NC2018_subset, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_NC2018_processed.csv")

write.csv(EPAair_PM25_NC2019_subset, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_NC2019_processed.csv")

```


## Combine datasets

7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code.

8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

  * Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels - but it will include sites with missing site information...) 

  * Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site name, AQS parameter, and county. Take the mean of the AQI value, latitude, and longitude.
  
  * Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
  
  * Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.

10. Call up the dimensions of your new tidy dataset.

11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1819_Processed.csv"

```{r}
#7 Combine the four datasets with `rbind'.
# Checking that column names are identical.
colnames(EPAair_O3_NC2018_subset)
colnames(EPAair_O3_NC2019_subset)
colnames(EPAair_PM25_NC2018_subset)
colnames(EPAair_PM25_NC2019_subset)

# Combining with 'rbind'.
EPAairData <- rbind(EPAair_O3_NC2018_subset, EPAair_O3_NC2019_subset, EPAair_PM25_NC2018_subset, EPAair_PM25_NC2019_subset)


#8 Create pipe for specified conditions. Dim = 14752 x 9.
EPAairData.Processed.1 <- 
  EPAairData %>%
  filter(Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School")) %>%
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  summarise(meanAQI = mean(DAILY_AQI_VALUE),
            meanLAT = mean(SITE_LATITUDE),
            meanLONG = mean(SITE_LONGITUDE)) %>%
  mutate(Month = month(Date), Year = year(Date))
  

dim(EPAairData.Processed.1)

EPAairData.Processed.FUCKALL <-
  EPAairData.Processed.1 %>%
   pivot_wider(names_from = "AQS_PARAMETER_DESC", values_from = "DAILY_AQI_VALUE")

# New pipe for second part
EPAairData.Processed.FUCKALL.TWO <- 
  EPAairData %>%
  filter(Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School")) %>%
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY, DAILY_AQI_VALUE) %>%
  mutate(meanLAT = mean(SITE_LATITUDE),
        meanLONG = mean(SITE_LONGITUDE),
        across(Row)) %>%
  mutate(Month = month(Date), Year = year(Date)) %>%
  pivot_wider(names_from = "AQS_PARAMETER_DESC", values_from = "DAILY_AQI_VALUE")

df %>%
  group_by(team) %>%
  mutate(mean_pts = mean(points)) %>%
  ungroup()

# Create new sequential row numbers for new dataset (as in dataset key).
EPAairData.Processed.TEST2 <- 
  EPAairData.Processed.1 %>%
  ungroup() %>%
  dplyr::mutate(New_Row=row_number()) %>%
  relocate(New_Row)
  
# Trying to work from processed 1, but don't think i can bc of row numbers and AQI values are "gone". We need row numbers to prevent duplicates and we need AQI values to pivot wider. But when i keep them, my dataset doesn't look like the key.

EPAairData.Processed.2 <-
  EPAairData.Processed.1 %>%
  



EPAairData.Processed.SPREAD <- 
  EPAairData %>%
  filter(Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School")) %>%
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  summarise(meanAQI = mean(DAILY_AQI_VALUE),
            meanLAT = mean(SITE_LATITUDE),
            meanLONG = mean(SITE_LONGITUDE)) %>%
  mutate(Month = month(Date), Year = year(Date)) %>%
  replace(EPAairData == "NULL", NA)
  

#9 Spreading dataset so that AQI values for ozone and PM2.5 are in separate columns. Retaining "Row" column so that there are no duplicate entries in PM2.5 column because in former attempts, this created lists instead of integer columns with multiple entries in cells. Unsure if we were supposed to complete this step directly from the last one (i.e. from the grouped and summarized dataset, but I was unable to complete the steps below from that point. I need to start with the orginal dataset first.)


EPAairData.Processed.SPREAD <- 
  EPAairData %>%
  filter(Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School")) %>%
  pivot_wider(names_from = "AQS_PARAMETER_DESC", values_from = "DAILY_AQI_VALUE")
  
EPAairData.Processed.SPREADFINAL <- 
  EPAairData.Processed.SPREAD %>%
  group_by(Date, Site.Name, COUNTY) %>%
  summarise(meanLAT = mean(SITE_LATITUDE),
            meanLONG = mean(SITE_LONGITUDE),
            across()) %>%
  mutate(Month = month(Date), Year = year(Date))
  
  
EPAairData.Processed.NONWORKING$Ozone[EPAairData.Processed.NONWORKING$Ozone == 'NULL'] <- NA 

EPAairData.Processed.NONWORKING$PM2.5[EPAairData.Processed.NONWORKING$PM2.5 == 'NULL'] <- NA 
  
  
  df[df == 'NULL'] <- NA

  ungroup() %>%
  dplyr::mutate(New_Row=row_number()) %>%
  relocate(New_Row)

  #trying to get last two columns from NULL to NA
  # Also, how do I keep the Row column without grouping by it, and without losing it?
  




EPAairData.Final.Spread <- 
  EPAairData.Processed %>%
  pivot_wider(names_from = "AQS_PARAMETER_DESC", values_from = "DAILY_AQI_VALUE")
  

#10 Finding dimensions of new dataset.
dim(EPAairData.Final.Spread)

#11 Saving processed dataset. During the spreading process, the Ozone and PM2.5 columns
# were converted into lists. So, first converting them back to integers. *NOTE* -- after creating row names for each of the datasets, this problem was eliminated and the file was able to be saved without issue.

write.csv(EPAairData.Final.Spread, row.names = FALSE, file = "./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv")

```


## Generate summary tables

12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where mean **ozone** values are not available (use the function `drop_na` in your pipe). It's ok to have missing mean PM2.5 values in this result. 

13. Call up the dimensions of the summary dataset.

```{r}

#12 Tried using the drop_na function, but couldn't get it to work anywhere (both figures just said "NA". The na.rm resulted in actual numbers (not NA output) but not sure if it's accurate.
EPAairData.Summary.2 <-
  EPAairData.Final.Spread %>%
  group_by(Site.Name, Month, Year) %>%
  summarize(meanOzone = mean(Ozone, na.rm = TRUE),
            meanPM2.5 = mean(PM2.5, na.rm = TRUE))

# OR DO YOU START FROM SCRATCH?? THEY DON't say!

            
 #13

dim(EPAairData.Summary.2)
```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: 
